Comparison among doublet predictions on benchmarking datasets

### 2024-06-03 analyses/doublet-detection/template-notebooks/03_compare-benchmark-results.Rmd
PCA section for comparison notebook

Import shared functions:
```{r}

plot_pca <- function(df, 
                     color_column, 
                     pred_column = NULL,
                     color_type = "d",
                     color_lab) {
  # Plot PCs colored by either discrete or continuous color variable, showing doublets on top
  # df is expected to contain columns PC1, PC2, `color_column`, and if color_type == "c" also `pred_column`. These should _not_ be provided as strings
  # By default, this function assumes a discrete color variable (`color_type = "d"`) with two levels. Use `color_type = "c"` for continuous.

  p <- ggplot(df) + 
    aes(x = PC1, 
        y = PC2, 
        color = {{color_column}}) +
  geom_point(alpha = 0.5) + 
  labs(color = color_lab) + 
  theme(
    legend.title.position = "top",
    legend.position = "bottom"
  )

  # Set the palette, and ensure doublets are on top for visibility
  if (color_type == "d") {
    p <- p + 
      scale_color_manual(values = c("black", "lightblue")) + 
    geom_point(
      data = dplyr::filter(df, {{color_column}} == "doublet"), 
      color = "black"
    )
  } else if (color_type == "c") {
    p <- p + 
      scale_color_viridis_c(direction = -1) + 
    geom_point(
      data = dplyr::filter(df, {{pred_column}} == "doublet"), 
      aes(color = {{color_column}})
    )
  }

  return(p)
}
```

@@

```{r read_data}
# Read in and data for analysis
doublet_df_list <- dataset_names |>
  purrr::map(
    \(dataset) {

@@

      # Add a consensus call column
      dataset_df <- dataset_df |>
        dplyr::rowwise() |>
        dplyr::mutate(consensus_call = dplyr::if_else(
          all(
            c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"
          ),
          "doublet", 
          "singlet"
        )) |>
        dplyr::mutate(
          call_type = dplyr::case_when(
            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
          ), 
          # set associated plotting colors
          call_type_color = dplyr::case_when(
            call_type == "tp" ~ "blue4",
            call_type == "tn" ~ "blue",
            call_type == "fp" ~ "brown1",
            call_type == "fn" ~ "darkorange1"
          )
          )

@@

## Viz: PCA

```{r, fig.width = 10}
doublet_df_list |>
  purrr::iwalk(
    \(df, dataset) {

      # First, ground truth
      p1 <- plot_pca(
        df, 
        color_column = ground_truth, 
        color_lab = "Ground truth"
      )

      # Second, consensus call
      p2 <- plot_pca(
        df, 
        color_column = consensus_call, 
        color_lab = "Consensus call"
      )

      # Third, call type
      p3 <- ggplot(df) + 
        aes(x = PC1, y = PC2, color = call_type_color) + 
        geom_point(alpha = 0.6, size = 0.25) + 
        scale_color_identity()

      # combine and plot
      plot( p1 + p2 + p3 + plot_annotation(glue::glue("PCA for {dataset}")) )
    }
  )
```

>>>>

```{r}
plot_pca_calls <- function(df, 
                           color_column, 
                           pred_column,
                           color_lab) {
  # Plot PCs colored by singlet/doublet, showing doublets on top
  # df is expected to contain columns PC1, PC2, `color_column`, and `pred_column`. These should _not_ be provided as strings
  ggplot(df) + 
    aes(x = PC1, 
        y = PC2, 
        color = {{color_column}}) +
  geom_point(
    size = 0.75, 
    alpha = 0.6
  ) +
  scale_color_manual(name = color_lab, values = c("black", "lightblue")) + 
  geom_point(
    data = dplyr::filter(df, {{color_column}} == "doublet"), 
    color = "black",
    size = 0.75
  ) +
  theme(
    legend.title.position = "top",
    legend.position = "bottom"
  )
}

plot_pca_metrics <- function(df, color_column, false_colors) {
  # Plot PCs colored by performance metric, showing false calls on top
  # false_colors is a vector of colors used for fn and fp points
  # df is expected to contain columns PC1, PC2, and `color_column`. This should _not_ be provided as a string.
  ggplot(df) + 
    aes(x = PC1, 
        y = PC2, 
        color = {{color_column}}) +
  geom_point(
    size = 0.75, 
    alpha = 0.6
  ) + 
  geom_point(
    data = dplyr::filter(df, {{color_column}} %in% false_colors), 
    size = 0.75
  ) +
  scale_color_identity() +
  theme(legend.position = "none")
}
```

@@

```{r read_data}
confusion_colors <- c(
  "tp" = "lightblue",
  "tn" = "pink",
  "fp" = "blue",
  "fn" = "firebrick2"
)

# Read in and data for analysis
doublet_df_list <- dataset_names |>
  purrr::map(
    \(dataset) {

@@

      # Add a consensus call column
      dataset_df <- dataset_df |>
        dplyr::rowwise() |>
        dplyr::mutate(consensus_call = dplyr::if_else(
          all(
            c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"
          ),
          "doublet", 
          "singlet"
        )) |>
        dplyr::mutate(
          call_type = dplyr::case_when(
            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
          ), 
          # set associated plotting colors
          call_type_color = dplyr::case_when(
            call_type == "tp" ~ unname(confusion_colors["tp"]),
            call_type == "tn" ~ unname(confusion_colors["tn"]),
            call_type == "fp" ~ unname(confusion_colors["fp"]),
            call_type == "fn" ~ unname(confusion_colors["fn"])
          )
          )

@@

## Visualizations

### PCA

This section plots the PCA for each dataset, with three color schemes from left to right:
- Ground truth doublets are shown in black
- Consensus doublets are shown in black
- Points are colored as tp, tn, fp, or fn based on comparing the consensus call to the ground truth

```{r, fig.width = 12, fig.height = 6}
# Make a legend for the confusion-colored PCA
legend_plot <- data.frame(
  x = factor(names(confusion_colors), levels = names(confusion_colors)), y = 1:4
) |>
 ggplot(aes(x = x, y = y, color = x)) + 
  geom_point(size = 3) + 
  scale_color_manual(name = "Metric", values = confusion_colors) 
confusion_legend <- ggpubr::get_legend(legend_plot) |> ggpubr::as_ggplot()

doublet_df_list |>
  purrr::iwalk(
    \(df, dataset) {

      # First, ground truth
      p1 <- plot_pca_calls(
        df, 
        color_column = ground_truth, 
        color_lab = "Ground truth"
      )

      # Second, consensus call
      p2 <- plot_pca_calls(
        df, 
        color_column = consensus_call, 
        color_lab = "Consensus call"
      )

      # Third, call type
      p3 <- plot_pca_metrics(
        df,
        call_type_color,
        false_colors = unname(c(confusion_colors["fn"], confusion_colors["fp"]))
      )

      # combine and plot
      plot( p1 + p2 + p3 + confusion_legend + plot_annotation(glue::glue("PCA for {dataset}")) + plot_layout(ncol=4, widths = c(1,1,1,0.25)) )
    }
  )
```
